---
title: "Weather and Income's impact on Capital Bikeshare Rides"
author: "Soren Snyder and Tahmina Alam"
date: "April 22, 2026"
format: 
  revealjs: 
    theme: moon
---


```{r}
#| label: setup
#| echo: false
#| output: false
#| cache: true

library(here)
library(tidyverse)

source(here("code", "00_load_packages.R"))
source(here("code", "04_combine_data.R"))
source(here("code", "05_regression_model.R"))

```



# Background

## Importance

- explain why this matters
- easy transportation should be accessible to those who need it the most
- does weather disproportionately affect lower income groups

## Research Question

write the research question

## Data

- Capital Bikeshare Data
- NOAA
  - DCA Weather Station
- Census (_need to be more exact__)


## Data Structure

```{r}
#| label: head of data

ready_data %>% 
  head() %>% 
  select(start_station_name, date, rides, est_med_income, prcp, snow, tmean) %>%
  kable(digits = 2) %>%
  kable_styling(
    full_width = FALSE
  ) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:2, color = "white")


```



## Challenges and Limitations with the Data


## Daily Rides

needs to change

```{r}
#| label: exploratory histogram
#| cache: true

par(mfrow = c(1, 1), bg = "azure3")
hist(model_data$rides, breaks = 50,
     col = "chocolate3",
     main = "Number of Rides per Day", col.main = "grey0",
     xlab = "Rides",
     ylab = "Number of Days",
     border = NA)

```

## Time Series

seasonality brings temperature as a potential predictor


# Modeling

## Goals

- interatiction term, count data

## Poisson

explain Poisson model..maybe use Latex to write out on another slide

## Negative Binomial

explain overdispersion of poisson model, so needed to swtich to negative binomial

## Comparison

```{r}
#| label: comparison table

model_compare %>%
  kable(digits = 2) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:2, color = "white")


```

# Interpretation

## Interaction term

look at model results table

##

```{r}
#| label: interaction plot
#| cache: true
#| echo: false

pred_grid <- expand.grid(
  tmean = seq(20, 90, by = 1),
  income_10k = c(5, 10, 15),
  prcp = 0,
  snow = 0,
  awnd = mean(model_data$awnd),
  poverty_pct = mean(model_data$poverty_pct),
  weekday = factor("Wed", levels = levels(model_data$weekday), ordered = TRUE)
)

pred_grid$predicted <- predict(nb_simp, newdata = pred_grid, type = "response")

low <- pred_grid[pred_grid$income_10k == 5,]
mid <- pred_grid[pred_grid$income_10k == 10,]
high <- pred_grid[pred_grid$income_10k == 15,]

# plot
par(bg = "azure3", mar = c(5, 5, 4, 2))

plot(low$tmean, low$predicted,
     type = "l",
     col = "chocolate3",
     lwd = 2,
     ylim = c(0, max(pred_grid$predicted) * 1.1),
     main = "Predicted Daily Rides by Temperature and Neighborhood Income",
     xlab = "Mean Temperature (°F)",
     ylab = "Predicted Rides per Station")

lines(mid$tmean,  mid$predicted,  col = "chartreuse3", lwd = 2)
lines(high$tmean, high$predicted, col = "mediumslateblue",  lwd = 2)

legend("topleft",
       legend = c("$50k", "$100k", "$150k"),
       col = c("chocolate3", "chartreuse3", "mediumslateblue"),
       lwd = 2,
       title = "Median Income",
       bg = "azure3")

```

## Coefficient Plot

plot here

## Conclusions

some conclusions

# Thank you









