---
title: "STAT-616 Final Project Brief"
author: "Soren Snyder"
format: pdf
date: "2026-03-11"
---

```{r}
#| label: run packages
#| warning: false
#| message: false

library(here)
source(here("code", "00_load_packages.R"))

```

```{r}
#| eval: false
# run eval: true if you don't already have the data

source(here("code", "01_bikeshare_data.R"))
source(here("code", "02_weather_data.R"))
source(here("code", "03_income_data"))


```


```{r}
#| label: create full, joined data set
#| warning: false
#| message: false

source(here("code", "04_combine_data.R"))

```



# Data

We pulled data from the Capital Bikeshare website and aggregated it to the number of rides per day. We pulled weather data from the DCA weather station using the NOAA api. We used the singular weather station because DC is a small enough geographic area where the area is impacted similar-enough, where it shouldn't impact the analysis. First, we tried to gather data from the several DC weather stations using the same api, but only one weather station had less than half their data missing. Going to one, reliable station simplifies the joining of data as well as makes the weather data more reliable. Finally, we used the census api to get income, poverty, and population data. The station and income data were joined geographically using lat and long coordinates of the bikeshare stations, then adding weather data by date.

```{r}
#| label: data head

ready_data %>% 
  head() %>% 
  select(start_station_name, date, rides, est_med_income, prcp, snow, tmean) %>%
  flextable() %>% 
  autofit() %>% 
  fit_to_width(max_width = 6.5)

```


We realized that I accidentally removed dates that had no rides, which would unintentionally bias our model. We fixed it by going back and adding all the dates after aggregating rides. Dates with missing values where then replaced with 0, indicating that no bikeshare rides where taken from that station that day.



# Main findings

## Poisson Regression Model

```{r}
#| label: Poisson Regression

source(here("code", "05_regression_model.R"))

```
____ We first used a Poisson regression model to study how daily bikeshare rides are related to factors like temperature, rain, snow, wind, income, poverty, and weekday.The Poisson model assumes that the mean and variance are the same. To check this, we calculated the dispersion statistic, which came out to be 11.77. This is much larger than 1, showing that the data has much more variability than the Poisson model can handle.Because of this overdispersion, the Poisson model does not fit the data well and is not reliable for drawing conclusions, although it still helps as a basic starting point.


```{r}

summary(pois_model)

```


## Negative Binomal Model

---- To deal with the overdispersion in the Poisson model, we fitted a Negative Binomial regression model, which is better suited when the data has extra variability.For this model, the dispersion statistic was about 1.08, which is very close to 1. This shows that the Negative Binomial model is able to handle the variability in the data much better.We also compared the predictive performance of both models using RMSE and found that the RMSE values were almost the same. This suggests that both models produce similar average predictions.In addition, we compared the models using AIC. The Poisson model had a very high AIC, while the Negative Binomial model had a much lower AIC, indicating a significantly better fit.Overall, these results suggest that the Negative Binomial model is more appropriate for this data.




