---
title: "Weather and Income's impact on Capital Bikeshare Rides"
author: "Soren Snyder and Tahmina Alam"
date: "April 22, 2026"
format: 
  revealjs: 
    theme: moon
---

```{r}
#| label: packages
#| echo: false
#| output: false

library(here)
source(here("code", "00_load_packages.R"))

```

```{r}
#| label: cleaning and models
#| echo: false
#| output: false
#| cache: true

source(here("code", "04_combine_data.R"))
source(here("code", "05_regression_model.R"))

```



# Background

## Importance

- explain why this matters
- easy transportation should be accessible to those who need it the most
- does weather disproportionately affect lower income groups

## Research Question

write the research question

## Data

- Capital Bikeshare Data
- NOAA
  - DCA Weather Station
- Census (_need to be more exact__)


## Data Structure

```{r}
#| label: head of data

ready_data %>% 
  head() %>% 
  select(start_station_name, date, rides, est_med_income, prcp, snow, tmean) %>%
  rename(
    Station = "start_station_name",
    Date = "date",
    Rides = "rides",
    `Med. Income` = "est_med_income",
    `Rain (in.)` = "prcp",
    `Snow (in.)` = "snow",
    `Mean Temp. (°F)` = "tmean"
  ) %>% 
  kable(digits = 2) %>%
  kable_styling(
    full_width = FALSE,
    font_size = 25
  ) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:6, color = "white")


```



## Challenges and Limitations with the Data


## Daily Rides

needs to change

```{r}
#| label: exploratory histogram
#| cache: true

par(mfrow = c(1, 1), bg = "azure3")
hist(model_data$rides, breaks = 50,
     col = "chocolate3",
     main = "Number of Rides per Day", col.main = "grey0",
     xlab = "Rides",
     ylab = "Number of Days",
     border = NA)

```

## Time Series

```{r}
#| label: time series plot
#| cache: true

total_rides_per_day <- aggregate(rides ~ date, data = model_data, FUN = sum)

par(bg = "azure3")
plot(total_rides_per_day$date, total_rides_per_day$rides,
     type = "l",
     col = "chocolate3",
     main = "Total Daily Bikeshare Rides (2025)",
     xlab = "Date",
     ylab = "Total Rides",
     xaxt = "n")
axis.Date(1, at = seq(min(total_rides_per_day$date), 
                      max(total_rides_per_day$date), 
                      by = "month"),
          format = "%b")
lines(total_rides_per_day$date,
      predict(loess(rides ~ as.numeric(date), data = total_rides_per_day, span = 0.2)),
      col = "grey20",
      lwd = 2,
      lty = 2)
legend("topleft",
       legend = c("Daily rides", "Trend"),
       col = c("chocolate3", "grey20"),
       lty = c(1, 2),
       lwd = c(1, 2),
       bg = "azure3")



```



# Modeling

## Goals

- interatiction term, count data

## Poisson

explain Poisson model..maybe use Latex to write out on another slide

## Negative Binomial

explain overdispersion of poisson model, so needed to swtich to negative binomial

## Comparison

```{r}
#| label: comparison table
#| cache: true

model_data$pred_pois <- predict(pois_model, type = "response")
rmse_pois <- sqrt(mean((model_data$rides - model_data$pred_pois)^2))

model_data$pred_nb <- predict(nb_simp, type = "response")
rmse_nb <- sqrt(mean((model_data$rides - model_data$pred_nb)^2))

model_compare <- tibble(
  Model = c("Poisson", "Negative Binomial"),
  AIC = c(AIC(pois_model), AIC(nb_simp)),
  Dispersion = c(
    deviance(pois_model) / df.residual(pois_model),
    deviance(nb_simp) / df.residual(nb_simp)
  ),
  RMSE = c(rmse_pois, rmse_nb)
)


# table
model_compare %>%
  kable(digits = 2) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:2, color = "white")

```

# Interpretation

## Interaction term


```{r}
#| label: nb model results
#| cache: true

tidy(nb_simp, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    term = c("Intercept", "Temperature", "Precipitation", "Snow", 
             "Wind", "Income ($10k)", "Poverty %", "Weekday.L", 
             "Weekday.Q", "Weekday.C", "Weekday^4", "Weekday^5", 
             "Weekday^6", "Temp × Income"),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(term, estimate, conf.low, conf.high, p.value, sig) %>%
  kable(
    digits = 3,
    col.names = c("Term", "IRR", "CI Low", "CI High", "p-value", "")
  ) %>%
  kable_styling(font_size = 20, full_width = FALSE) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:14, color = "white") %>%
  row_spec(14, bold = TRUE)

```


##

```{r}
#| label: interaction plot
#| cache: true
#| echo: false

pred_grid <- expand.grid(
  tmean = seq(20, 90, by = 1),
  income_10k = c(5, 10, 15),
  prcp = 0,
  snow = 0,
  awnd = mean(model_data$awnd),
  poverty_pct = mean(model_data$poverty_pct),
  weekday = factor("Wed", levels = levels(model_data$weekday), ordered = TRUE)
)

pred_grid$predicted <- predict(nb_simp, newdata = pred_grid, type = "response")

low <- pred_grid[pred_grid$income_10k == 5,]
mid <- pred_grid[pred_grid$income_10k == 10,]
high <- pred_grid[pred_grid$income_10k == 15,]

# plot
par(bg = "azure3", mar = c(5, 5, 4, 2))

plot(low$tmean, low$predicted,
     type = "l",
     col = "chocolate3",
     lwd = 2,
     ylim = c(0, max(pred_grid$predicted) * 1.1),
     main = "Predicted Daily Rides by Temperature and Neighborhood Income",
     xlab = "Mean Temperature (°F)",
     ylab = "Predicted Rides per Station")

lines(mid$tmean,  mid$predicted,  col = "chartreuse3", lwd = 2)
lines(high$tmean, high$predicted, col = "mediumslateblue",  lwd = 2)

legend("topleft",
       legend = c("$50k", "$100k", "$150k"),
       col = c("chocolate3", "chartreuse3", "mediumslateblue"),
       lwd = 2,
       title = "Median Income",
       bg = "azure3")

```

## Coefficient Plot

```{r}
#| label: coefficient plot
#| cache: true

nb_coef <- tidy(nb_simp, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

ggplot(nb_coef, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  labs(
    title = "Incidence Rate Ratios from Negative Binomial Model",
    x = "Predictor",
    y = "IRR"
  )


```

## Conclusions

some conclusions

# Thank you









