---
title: "Weather and Income's impact on Capital Bikeshare Rides"
author: "Soren Snyder and Tahmina Alam"
date: "April 22, 2026"
format: 
  revealjs: 
    theme: moon
---

```{r}
#| label: packages
#| echo: false
#| output: false

library(here)
source(here("code", "00_load_packages.R"))

# all of the code can be found on this GitHub Repo:
# https://github.com/ss6662a/STAT-616_FP

```

```{r}
#| label: cleaning and models
#| echo: false
#| output: false
#| cache: true

source(here("code", "04_combine_data.R"))
source(here("code", "05_regression_model.R"))

```



# Background

## Importance

- Bikesharing is a popular, low-cost method of transportation in cities
- Transportation is critical 
- Easy and cheap transportation methods should be accessible to those who need it the most
- If weather disproportionately affects lower income neighborhoods, transportation eauity could be affected

## Research Question {.center}

Does weather have a more extreme effect on Capital Bikeshare ridership in lower income neighborhoods in DC?

## Data

- Capital Bikeshare Data
- National Oceanic and Atmospheric Administration (NOAA)
- Census Income Data

```{r}
#| label: summary stats

data.frame(
  Variable = c("Rides", "Mean Temp (°F)", "Precipitation (in.)", 
               "Snow (in.)", "Wind Speed", "Income ($10k)", "Poverty %"),
  N = format(nrow(model_data), big.mark = ","),
  Min = c(min(model_data$rides), min(model_data$tmean), min(model_data$prcp),
          min(model_data$snow), min(model_data$awnd), min(model_data$income_10k),
          min(model_data$poverty_pct)),
  Mean = c(mean(model_data$rides), mean(model_data$tmean), mean(model_data$prcp),
           mean(model_data$snow), mean(model_data$awnd), mean(model_data$income_10k),
           mean(model_data$poverty_pct)),
  Max = c(max(model_data$rides), max(model_data$tmean), max(model_data$prcp),
          max(model_data$snow), max(model_data$awnd), max(model_data$income_10k),
          max(model_data$poverty_pct))
) %>%
  kable(digits = 2) %>%
  kable_styling(font_size = 20, full_width = FALSE, position = "left") %>%
  row_spec(0, color = "white") %>%
  row_spec(1:7, color = "white") %>% 
  footnote(general = "392 stations",
           general_title = "",
           footnote_as_chunk = TRUE)

```


## Challenges and Limitations with the Data

- Missing values
  - Removing e-bikes
  - DCA weather station
  - Imputation for wind speed
- Rides outside DC boundaries were removed


## Data Structure

```{r}
#| label: head of data

ready_data %>% 
  head() %>% 
  select(start_station_name, date, rides, est_med_income, prcp, snow, tmean) %>%
  rename(
    Station = "start_station_name",
    Date = "date",
    Rides = "rides",
    `Med. Income` = "est_med_income",
    `Rain (in.)` = "prcp",
    `Snow (in.)` = "snow",
    `Mean Temp. (°F)` = "tmean"
  ) %>% 
  kable(digits = 2) %>%
  kable_styling(
    full_width = FALSE,
    font_size = 25
  ) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:6, color = "white")


```




## Daily Rides

```{r}
#| label: exploratory histogram

par(mfrow = c(1, 1), bg = "azure3")
hist(model_data$rides, breaks = 50,
     col = "chocolate3",
     main = "Number of Rides per Day", col.main = "grey0",
     xlab = "Rides",
     ylab = "Number of Days",
     border = NA)

```

## Time Series

```{r}
#| label: time series plot

total_rides_per_day <- aggregate(rides ~ date, data = model_data, FUN = sum)

par(bg = "azure3")
plot(total_rides_per_day$date, total_rides_per_day$rides,
     type = "l",
     col = "chocolate3",
     main = "Total Daily Bikeshare Rides (2025)",
     xlab = "Date",
     ylab = "Total Rides",
     xaxt = "n")
axis.Date(1, at = seq(min(total_rides_per_day$date), 
                      max(total_rides_per_day$date), 
                      by = "month"),
          format = "%b")
lines(total_rides_per_day$date,
      predict(loess(rides ~ as.numeric(date), data = total_rides_per_day, span = 0.2)),
      col = "grey20",
      lwd = 2,
      lty = 2)
legend("topleft",
       legend = c("Daily rides", "Trend"),
       col = c("chocolate3", "grey20"),
       lty = c(1, 2),
       lwd = c(1, 2),
       bg = "azure3")



```



# Modeling

## Goals

- Want to model a daily ride count using weather, income, and other socioeconomic predictors
	- Poisson model
- Does the effect of weather vary by neighborhood income?
	- Weather x income interactions

## Poisson

- Modeling count data
- Predictors relate to ride counts on a log scale
- Coefficients are exponentiated to get Incidence Rate Ratios (IRRs)
- Key assumption: mean = variance


## Poisson Results

- Poisson assumption (mean = variance) was badly violated
- Poisson model dispersion statistic = `r round(poi_overdispersion, 2)`
	- Standard errors become too small, making results look more significant than they are
- Negative Binomial model adds a parameter to absorb extra variability
- Dispersion drops to `r round(nb_overdispersion, 2)`


## Negative Binomial 
### Model

- First: full model with all weather × income interactions
  - Only temperature × income was significant
  - Precipitation, snow, and wind interactions were not
- Then fitted a simplified model
  - AIC improved slightly

## Comparison

- AIC drops dramatically with Negative Binomial
- Dispersion is much better on NB
- RMSE very similar

```{r}
#| label: comparison table
#| cache: true

model_data$pred_pois <- predict(pois_model, type = "response")
rmse_pois <- sqrt(mean((model_data$rides - model_data$pred_pois)^2))

model_data$pred_nb <- predict(nb_simp, type = "response")
rmse_nb <- sqrt(mean((model_data$rides - model_data$pred_nb)^2))

model_compare <- tibble(
  Model = c("Poisson", "Negative Binomial"),
  AIC = c(AIC(pois_model), AIC(nb_simp)),
  Dispersion = c(
    deviance(pois_model) / df.residual(pois_model),
    deviance(nb_simp) / df.residual(nb_simp)
  ),
  RMSE = c(rmse_pois, rmse_nb)
)


# table
model_compare %>%
  kable(digits = 2) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:2, color = "white")

```

# Interpretation

## Model Results

```{r}
#| label: nb model results
#| cache: true

tidy(nb_simp, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    term = c("Intercept", "Temperature", "Precipitation", "Snow", 
             "Wind", "Income ($10k)", "Poverty %", "Weekday.L", 
             "Weekday.Q", "Weekday.C", "Weekday^4", "Weekday^5", 
             "Weekday^6", "Temp × Income"),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(term, estimate, std.error, conf.low, conf.high, p.value, sig) %>%
  kable(
    digits = 3,
    col.names = c("Term", "IRR", "SE", "CI Low", "CI High", "p-value", "")
  ) %>%
  kable_styling(font_size = 20, full_width = FALSE) %>%
  row_spec(0, color = "white") %>%
  row_spec(1:14, color = "white") %>%
  row_spec(14, bold = TRUE)

```


## Temperature x Income Interaction


```{r}
#| label: interaction plot

pred_grid <- expand.grid(
  tmean = seq(20, 90, by = 1),
  income_10k = c(5, 10, 15),
  prcp = 0,
  snow = 0,
  awnd = mean(model_data$awnd),
  poverty_pct = mean(model_data$poverty_pct),
  weekday = factor("Wed", levels = levels(model_data$weekday), ordered = TRUE)
)

pred_grid$predicted <- predict(nb_simp, newdata = pred_grid, type = "response")

low <- pred_grid[pred_grid$income_10k == 5,]
mid <- pred_grid[pred_grid$income_10k == 10,]
high <- pred_grid[pred_grid$income_10k == 15,]

# plot
par(bg = "azure3", mar = c(5, 5, 4, 2))

plot(low$tmean, low$predicted,
     type = "l",
     col = "chocolate3",
     lwd = 2,
     ylim = c(0, max(pred_grid$predicted) * 1.1),
     main = "Predicted Daily Rides by Temperature and Neighborhood Income",
     xlab = "Mean Temperature (°F)",
     ylab = "Predicted Rides per Station")

lines(mid$tmean,  mid$predicted,  col = "chartreuse3", lwd = 2)
lines(high$tmean, high$predicted, col = "mediumslateblue",  lwd = 2)

legend("topleft",
       legend = c("$50k", "$100k", "$150k"),
       col = c("chocolate3", "chartreuse3", "mediumslateblue"),
       lwd = 2,
       title = "Median Income",
       bg = "azure3")

```



## Coefficient Plot

```{r}
#| label: coefficient plot
#| cache: true

nb_coef <- tidy(nb_simp, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = c("Temperature", "Precipitation", "Snow", "Wind", 
                  "Income ($10k)", "Poverty %", "Weekday.L", "Weekday.Q", 
                  "Weekday.C", "Weekday^4", "Weekday^5", "Weekday^6", 
                  "Temp × Income"))

par(bg = "azure3", mar = c(5, 8, 4, 2))

plot(nb_coef$estimate, 1:nrow(nb_coef),
     xlim = c(min(nb_coef$conf.low) * 0.9, max(nb_coef$conf.high) * 1.1),
     yaxt = "n",
     pch = 19,
     col = "chocolate3",
     xlab = "Incidence Rate Ratio (IRR)",
     ylab = "",
     main = "Incidence Rate Ratios — Negative Binomial Model")

segments(nb_coef$conf.low, 1:nrow(nb_coef),
         nb_coef$conf.high, 1:nrow(nb_coef),
         col = "chocolate3", lwd = 2)

abline(v = 1, lty = 2, col = "grey20", lwd = 1.5)

axis(2, at = 1:nrow(nb_coef),
     labels = nb_coef$term,
     las = 2,
     cex.axis = 0.85)


points(nb_coef$estimate[nb_coef$term == "Temp × Income"],
       which(nb_coef$term == "Temp × Income"),
       pch = 19, col = "chocolate4", cex = 1.5)


```

## Rides by Day of Week

```{r}
#| label: rides by day of week

weekday_grid <- expand.grid(
  weekday = factor(levels(model_data$weekday), 
                   levels = levels(model_data$weekday), 
                   ordered = TRUE),
  tmean = mean(model_data$tmean, na.rm = TRUE),
  income_10k  = mean(model_data$income_10k, na.rm = TRUE),
  prcp = 0,
  snow = 0,
  awnd = mean(model_data$awnd, na.rm = TRUE),
  poverty_pct = mean(model_data$poverty_pct, na.rm = TRUE)
)

weekday_grid$predicted <- predict(nb_simp, newdata = weekday_grid, type = "response")

par(bg = "azure3", mar = c(5, 5, 4, 2))

plot(1:7, weekday_grid$predicted,
     type = "b",
     pch = 19,
     col = "chocolate3",
     lwd = 2,
     xaxt = "n",
     main = "Predicted Daily Rides by Day of Week",
     xlab = "Day of Week",
     ylab = "Predicted Rides per Station")

axis(1, at = 1:7, labels = levels(model_data$weekday))

```


## Conclusions

- Weather affects ridership broadly
- The temperature x income interaction is minimal but significant
- Precipitation and snow suppress ridership uniformly regardless of income
- Income independently predicts ridership, suggesting access gaps beyond just weather sensitivity

# Thank you









