---
title: "STAT-616 Final Project Brief"
author: "Soren Snyder"
format: pdf
date: "2026-03-11"
---

```{r}
#| label: run packages
#| warning: false
#| message: false

library(here)
source(here("code", "00_load_packages.R"))

```

```{r}
#| eval: false
# run eval: true if you don't already have the data

source(here("code", "01_bikeshare_data.R"))
source(here("code", "02_weather_data.R"))
source(here("code", "03_income_data"))


```


```{r}
#| label: create full, joined data set
#| warning: false
#| message: false

source(here("code", "04_combine_data.R"))

```



# Data

We pulled data from the Capital Bikeshare website and aggregated it to the number of rides per day. We pulled weather data from the DCA weather station using the NOAA api. We used the singular weather station because DC is a small enough geographic area where the area is impacted similar-enough, where it shouldn't impact the analysis. First, we tried to gather data from the several DC weather stations using the same api, but only one weather station had less than half their data missing. Going to one, reliable station simplifies the joining of data as well as makes the weather data more reliable. Finally, we used the census api to get income, poverty, and population data. The station and income data were joined geographically using lat and long coordinates of the bikeshare stations, then adding weather data by date.

```{r}
#| label: data head

ready_data %>% 
  head() %>% 
  select(start_station_name, date, rides, est_med_income, prcp, snow, tmean) %>%
  flextable() %>% 
  autofit() %>% 
  fit_to_width(max_width = 6.5)

```


We realized that I accidentally removed dates that had no rides, which would unintentionally bias our model. We fixed it by going back and adding all the dates after aggregating rides. Dates with missing values where then replaced with 0, indicating that no bikeshare rides where taken from that station that day.



# Main findings

## Poisson Regression Model

```{r}
#| lable: Poisson Regression

source(here("code", "05_regression_model.R"))

```



__ talk about how the deviance is very high because of overdispersion


## Negative Binomal Model



In order to __ the high deviance and variance from the Poisson model, we tried to model the data using the negative binomial model

highlight the AIC and variance/deviance and dispersion (poisson 11.77 vs bi nominal _lower_)


----

found that the rmse for each was the same


